\large How resource abundance and stochasticity affect animals' spatial needs

### 2022-11-17 writing/manuscript.Rmd
cleaned sections; added refs

@@

@@

While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce. Two years later, @harestad_home_1979 proposed that the simplest hypothesis for the relationship between an animal's home range size is inversely proportional to resource abundance. More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022]. Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020]. Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists [@lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982]. Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs. **but animals don't only respond to changes in mean**

**paragraph detailing effects of variance on movement and how little work has been done**

@@

In this paper, we can simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$ (which we write as $R \sim \Gamma(\mu(t), \sigma^2(t))$). Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ to facilitate visualizing the examples. However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean. As the mean approaches zero (from the positive side, which we write as $\mu \rightarrow0^+$) the variance also does: $\mu \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \sigma^2 = 0^+$.

@@

The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019]. Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the positional variance of animals that are not range-resident.

@@

Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement. Animals may temporarily change their behavior in response to many causes, including predators [@ref], competitors [@ref], weather [@ref], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].

@@

@@

<!-- You need to open with some background on metabolic requirements. Individuals have fixed needs that they must meet in order to survive, grow, and reproduce. If these are not met, they will die. Then transition to the idea that, because of that, they move around in order to secure resources. Therefore, when environments are productive, they don't have to move as far to meet their energetic requirements. -->

* add refs to generalize beyond terrestrial mammals

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@baldwin_nutritional_1984; @reich_body_2001; @ref]. Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism [@ref] and movement costs [@ref]. The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @ex]. Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp). As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].

@@

\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], $\var(R)$ has received far less attention [but see @lucherini_habitat_1996; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021]. However, $\var(R)$ can change strongly due to many important causes, including repetitive and predictable patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches) as well as infrequent and unpredictable events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018]. Environmental variability, including extreme events [@logares_black_2012], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness and increases their spatial needs. A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited. There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.

Simlarly to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on $\vpos$. When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect on $\vpos$. However, as resources become additionally unpredictable, we expect spatial needs to increase superlinearly. If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors, including nomadism, dispersal, and migration (figure \@ref(fig:s2-hyp)).

The effects of $\var(R)$ on animals' spatial requirements most likely vary between species, since different species have different energetic and nutritional needs, so the degree to which $R$ varies will likely affects each species differently. Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with a larger $\vpos$ may not detect such differences. @rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity. Similarly, @mueller_how_2011 demonstrated that ungulate species in more variable ecosystems move more than species in more stable landscapes and that they are more likely to be nomadic. In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous habitats) required exponentially less space than foxes that lived in more homogeneous regions. Similarly, @nilsen_can_2005 found that the home ranges of wolverines (*Gulo gulo*) and Canadian lynx (*Lynx canadensis*) decreased with seasonality, while the home ranges of bobcats (*Lynx rufus*) and coyotes (*Canis latrans*) increased with seasonality. The work by @nilsen_can_2005 also demonstrates that the effect of $\e(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\var(R)$, so it is important to account for interaction effects between the two.

* predation is more stochastic than herbibory; predators need to match more cycles, depend on fewer food sources that are able to move $\rightarrow$ higher effect of $\var(R)$

@@

\noindent Consider the simplistic scenario where $\e(R) = \mu$ changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that the resource abundance follows a Gamma distribution with a mean that changes over time (i.e., the expected value is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d). However, if $\e(R) = \mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019]. However, since $\e(R)$ does not vary predictably over space, it will not be possible to predict the location of locations with high or low $R$.

This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions with little to no changes in variability and where unpredictable events occur at a similar frequency over time, or when data availability is too low to produce appreciable measures of variance. Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case. Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response. Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' $\vpos$ should be estimated before assuming the effect is negligible. The need to account for $\var(R)$ is compounded by recent changes in climate, which exposes species to novel situations and increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018]. Additionally, anthropogenic structures and urban areas reduce the habitat available to many terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018]. Although some species rely on anthropogenic areas and structures for predicable and dependable sources of resources [steppe rat snake: @yu_artificial_2022; yellow mongoose: @cronk_home_2021; common octopus: @arechavala-lopez_common_2019; raccoon dog: @mitsuhashi_home_2018; coyote: @peron_periodic_2017], anthropogenic environments can be highly unpredictable [@ref] and pose high mortality risks [@noonan_roads_2021; @refs].

@@

Most habitats with seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of dormancy or lower productivity, and the variability in $R$ often changes both within and between years. In such cases, failing to account for $\var(R)$ can result in serious biases in the estimation of animals' spatial needs [@examples]. In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or in periods of increased energetic needs, such as reproductive periods, whether this be due to pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [@male_deer_growing_antlers]. In heterogeneous environments, such as habitats with a high diversity of clustered resources, animals may have to move between locations if they cannot depend on a single one.

# Interactive effects of mean and variance in resource abundance

Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$. The top row (constant $\var(R)$) shows how $\vpos$ changes as a function of $\e(R)$ while $\var(R)$ remains constant. (Some noise in $H$ is still present since $\var(R) \ne 0$ and thus $R$ is not constant). Note the nonlinear decrease in $H$ as $\e(R)$ increases at a constant slope (linear $\e(R)$), with larger changes when $\e(H)$ is low. This is because when $\e(R)$ is low, small additive changes in $\e(R)$ (e.g. from 1 to 2) can correspond to large multiplicative changes (i.e., 2 is twice as big as 1). Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$, since (not sure about this) larger quantiles (e.g., 95% > 50%) are more sensitive to changes in a distribution's variance (in this case $\propto \vpos$).

Most habitats with seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of "dormancy" or lower productivity, and the variability in $R$ often changes both within and between years. In such cases, failing to account for $\var(R)$ can result in serious biases in the estimation of animals' spatial needs.

\noindent Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$. The top row (constant $\var(R)$) shows how $\vpos$ remains constant. Some noise in $H$ is still present since $\var(R) \ne 0$ and thus $R$ is not constant). Note the nonlinear decrease in $H$ when $\e(R)$ increases at a constant slope (linear $\e(R)$), with larger changes when $\e(R)$ is low. This is because when $\e(R)$ is low, small additive changes in $\e(R)$ (e.g. from 1 to 2) can correspond to large multiplicative changes (i.e., 2 is twice as big as 1). Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$ since larger quantiles (e.g., 95% > 50%) are more sensitive to changes in the $\vpos$. In regions where the average resource abundance changes over time (cyclical $\e(R)$; e.g., seasonal changes), animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\e(R)$ (e.g., winter park closures). However, estimates of spatial requirements based on estimated changes in $\e(R)$ should be interpreted carefully, since model error and unpredictable changes in $\e(R)$ may increase animals' spatial needs unpredictably. Thus, it is best to avoid changing the space available to animals too close to when resources are expected to be low and include a "buffer" area so the available area is larger than the estimated spatial needs. This is particularly the case in environments where resource abundance changes unpredictably (drifting $\e(R)$), since accurate long-term estimates of $\e(R)$ may be hard to produce, if not impossible. In cases where $\e(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time. Finally, in ecosystems where $\e(R)$ changes erratically and suddenly (erratic $\e(R)$), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.

@@

Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios. $\e(R)$ and $\var(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\e(R)$ and $\var(R)$ remain undetected. Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly continuously, such increases may be possible for short periods of time (followed by periods of no change or decrease). Additionally, these examples are important because they demonstrate the relationships between $H$, $\e(R)$, and $\var(R)$ in (relatively) simple scenarios. Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity. Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation). However, if highly stochastic or erratic changes occur frequently, animals may perceive them as smooth changes in $\e(R)$ or increases in $\var(R)$ rather a series of small, sudden, changes. Estimating the true trend in $\e(R)$ and $\var(R)$ would often require an excessively high cognitive capacity and an equally unlikely abundance of information, since although changes in $\e(R)$ are not due to $\var(R)$, but often distinguishing between the two is not easy [@steixner-kumar_strategies_2020], and $\e(R)$ and $\var(R)$ are not independent for $R \sim \Gamma(\mu, \sigma^2)$.

@@

>>>>

\singlespacing

@@

\doublespacing

@@

While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce. Two years later, @harestad_home_1979 proposed that the simplest hypothesis for the relationship between an animal's home range size is inversely proportional to resource abundance. More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022]. Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020]. Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982]. Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs. And while abundant work has been done on estimating animal's response to changes in resource abundance, there is little to no information on how many species respond to unpredictable changes in the resources they depend on.

A habitat's long-term average resource abundance is not sufficient to assess the habitat's quality. A grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with rare, sudden, and strong pulses of resources. In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food, while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are exhausted. Although both habitats may have the same long-term resource abundance, the differences in resource predictability result in substantially different habitats. Despite this, little work has been done on estimating the effects of resource unpredictability on animals' spatial needs [but see @rizzuto_forage_2021; @herfindal_prey_2005; @nilsen_can_2005].

@@

In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$ (which we write as $R \sim \Gamma(\mu(t), \sigma^2(t))$). Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ to facilitate visualizing the examples. However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean. As the mean approaches zero (from the positive side, which we write as $\mu \rightarrow0^+$) the variance also does: $\mu \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \sigma^2 = 0^+$.

@@

The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019]. Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the positional variance of animals that are not range-resident. ***but we can use a sliding window approach if animals are temporarily range resident or to estimate continuous changes in short-term spatial needs***

@@

Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement. Animals may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].

@@

@pretorius_movement_2020: flamingos -- migrating or nomadic?

@@

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@harvey_primate_1981; @ref]. The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022]. Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_20129; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan (figure \@ref(fig:mu-hyp)). As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].

@@

\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], $\var(R)$ has received far less attention [but see @lucherini_habitat_1996; @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021]. However, $\var(R)$ can change strongly due to many important causes, including repetitive and predictable patterns as well as infrequent and unpredictable events. Examples of predictable changes in $\var(R)$ include changes that occur between seasons, such as changes in temperature in continental summers (e.g., between 15 and 30 $^\circ$C) and winter temperatures (e.g., from -40 to -10$^\circ$C within the span of a few days), changes in precipitation frequency and unpredictability between the dry season (predictable lack of rain) and the wet season (frequent but sudden and unpredictable rains), or changes in fruit and berry abundance with the arrival of spring and summer (since the timing between plants can depend on many unaccounted factors). Although each of these events is unpredictable, the change in unpredictability can be predicted and expected. In contrast, other sources of stochasticity can be much less predictable, such as fires, floods, droughts, the arrival of new competitors, and changes in climate [@ipcc_global_2018; @noonan_situ_2018].

Environmental variability, including extreme events [@logares_black_2012], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness and increases their spatial needs. A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited. There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.

Similarly to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs. When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)). If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage, torpor, and hibernation).

The effects of $\var(R)$ on animals' spatial requirements most likely vary between species, since different species have different energetic and nutritional needs [@boratynski_energetic_2020; @noonan_effects_2020], so the degree to which $R$ varies will likely affects each species differently. Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with larger spatial needs may not detect such differences. @rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity. Similarly, @mueller_how_2011 demonstrated that ungulate species in more variable ecosystems move more than species in more stable landscapes and that they are more likely to be nomadic. In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous habitats) required exponentially less space than foxes that lived in more homogeneous regions. Similarly, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality. A subsequent analysis by @nilsen_can_2005 found that various carnivores respond to resource variability and seasonality (albeit in contrasting ways) using the Canadian lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*). Additionally, @nilsen_can_2005 also demonstrate that the effect of $\var(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\e(R)$, so it is not only important to account for both $e(R)$ and $\var(R)$ but also interaction effects between the two. It is worth noting that carnivores may be more susceptible to $\var(R)$ since animal prey's ability to move makes their abundance and location less predictable than those of plants. Since, the effects of $e(R)$ and $\var(R)$ may also affect prey abundance, it is important to account for both $e(R)$ and $\var(R)$ as well as any interactive effects between the two and how the effects may vary between animals with different diets or from different trophic levels.

@@

\noindent Consider the simplistic scenario where the average in resource abundance, $\e(R) = \mu$, changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d). However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and decreases in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017; @abrahms_memory_2019]. However, since $\e(R)$ does not vary predictably over space, it will not be possible to predict the location of locations with high or low $R$.

This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions with little to no changes in variability and where unpredictable events occur at a similar frequency over time, or when data availability is too low to produce appreciable measures of variance. Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case. Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response. Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' spatial needs should be estimated before assuming the effect is negligible. The need to account for $\var(R)$ is compounded by recent changes in climate, which exposes species to novel situations and increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018]. Additionally, anthropogenic structures and urban areas reduce the habitat available to many terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018]. Although some species rely on anthropogenic areas and structures for predicable and dependable sources of resources [wild boar: @ikeda_home_2022; elephants: @benitez_precipitation_2022; steppe rat snake: @yu_artificial_2022; yellow mongoose: @cronk_home_2021; common octopus: @arechavala-lopez_common_2019; raccoon dog: @mitsuhashi_home_2018; coyote: @peron_periodic_2017], anthropogenic environments can be highly unpredictable and pose high mortality risks [giant anteaters: @noonan_roads_2021; killer whales: @matkin_ongoing_2008].

@@

Most habitats with seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of dormancy or lower productivity, and the variability in $R$ often changes both within and between years. In such cases, failing to account for $\var(R)$ can result in serious biases in the estimation of animals' spatial needs [@examples]. In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or in periods of increased energetic needs, such as during pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022]. In heterogeneous environments, such as habitats with a high diversity of clustered resources, animals may have to move between locations if they cannot depend on a single one [@lucherini_habitat_1996].

Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$. The top row (constant $\var(R)$) shows how $\vpos$ remains constant. Some noise in $H$ is still present since $\var(R) \ne 0$ and thus $R$ is not constant). Note the nonlinear decrease in $H$ when $\e(R)$ increases at a constant slope (linear $\e(R)$), with larger changes when $\e(R)$ is low. This is because when $\e(R)$ is low, small additive changes in $\e(R)$ (e.g. from 1 to 2) can correspond to large multiplicative changes (i.e., 2 is twice as big as 1). Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$ since larger quantiles (e.g., 95% > 50%) are more sensitive to changes in the $\vpos$. In regions where the average resource abundance changes over time (cyclical $\e(R)$; e.g., seasonal changes), animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\e(R)$ (e.g., winter park closures). However, estimates of spatial requirements based on estimated changes in $\e(R)$ should be interpreted carefully, since model error and unpredictable changes in $\e(R)$ may increase animals' spatial needs unpredictably. Thus, it is best to avoid changing the space available to animals too close to when resources are expected to be low and include a "buffer" area so the available area is larger than the estimated spatial needs. This is particularly the case in environments where resource abundance changes unpredictably (drifting $\e(R)$), since accurate long-term estimates of $\e(R)$ may be hard to produce, if not impossible. In cases where $\e(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time. Finally, in ecosystems where $\e(R)$ changes erratically and suddenly (erratic $\e(R)$), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.

@@

Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios. $\e(R)$ and $\var(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\e(R)$ and $\var(R)$ remain undetected. Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly continuously, such increases may be possible for short periods of time (followed by periods of no change or decrease). Additionally, these examples are important because they demonstrate the relationships between $H$, $\e(R)$, and $\var(R)$ in (relatively) simple scenarios. Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity. Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation). However, if highly stochastic or erratic changes occur frequently, animals may perceive them as smooth changes in $\e(R)$ or increases in $\var(R)$ rather a series of small, sudden, changes. Estimating the true trend in $\e(R)$ and $\var(R)$ would often require an excessively high cognitive capacity and an equally unlikely abundance of information, since although changes in $\e(R)$ are not due to $\var(R)$, but often distinguishing between the two is not easy [@steixner-kumar_strategies_2020], and $\e(R)$ and $\var(R)$ are not independent for $R \sim \Gamma(\mu, \sigma^2)$.

@@

@ranc_memory_2022: Roe deer create HR based on memory
